---
title: "ENVS 193 Final Script"
author: "Jared Umphress"
subtitle: "June 9, 2026"
format: pdf

execute:
  warning: false
  message: false
---

## Set Up\newline

[Link to GitHub repository](https://github.com/Jared363/ENVS-193DS_spring-2026_final)


Read in packages 
```{r}

library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(ggplot2)
library(ggeffects)
library(flextable)

```

Read in data
```{r}

# bringing in screen_time.csv (personal data) with read_csv function 
#   and storing it as object called my_data
my_data <- read_csv(here("data/screen_time.csv"))

# bringing in nest_data_final.csv with read_csv function and storing it 
#   as object called nest_data
nests <- read_csv(here("data/nest_data_final.csv"))

```

## Problem 1. Research writing

a. Transparent statistical methods\newline

In part 1 of the results section, they used a logistic regression. In part 2 of the results sections, they used a chi squared test.

b. Suggestions for rewriting\newline

Our results suggest that distance to water edge predicts the probability of American Avocet microhabitat use. In addition, we found evidence to support that there is a relationship between birds species (American Avocet, Foster's Tern, Black-necked Stilt) and habitat (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat).

c. Further analysis\newline

Algae concentration is another variable that could influence the probability of American Avocet microhabitat use as algae is another important primary producer in aquatic ecosystems that influences food availability. It is a continuous predictor variable measured in mg dry weight/L water, and it can be collected in 1 liter containers with the dry weight measured in lab. 

## Problem 2. Data analysis

a. Clean and wrangle\newline
```{r}

# cleaning nests data and storing it as nests_clean
nests_clean <- nests |>
  # selecting for just three columns
  select(height_cm, case_control, ant_species) |>
  # filtering for only three ant species
  filter(ant_species %in% c("RRB","AB","BBR")) |>
  # creating a new column to store full scientific names for ants
  #   based on abreviations
   mutate(species_name = case_match(
    ant_species,
    "RRB" ~ "Crematogaster mimosae", 
    "AB" ~ "Crematogaster sjostedti", 
    "BBR" ~ "Crematogaster nigriceps"
  )) |>
  # ordering the species level
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                       "AB",
                                       "RBR",
                                       "BBR"))

# display structure of nests_clean data
str(nests_clean)

# display 10 random rows from the nests_clean data
slice_sample(nests_clean, 
             n = 10, 
             by = NULL, 
             weight_by = NULL, 
             replace = FALSE)

```

b. Response variable\newline

For the nest occurence response variable, a 1 indicates that that a tree contained a nest, and 0 indicates that a tree was available for a nest. 

c. Purpose of study\newline

Tree height, one of the predictor variables, is a continuous variable. As tree height increases, the probability of nest occurence would increase because the nests are further off the ground and harder to get to for predators (Results, paragraph 2).

Ant species, the other predictor variable, is a categorical variable with the four ant species representing four categories. The presence of *C. nigriceps* and *C. mimosae* would increase nest occurence as they are aggressive defenders of host trees and thus can reduce the risk of nest predation compared to *C. sjostedi* and *T. penzigi* which are less aggressive host tree defenders (Discussion, paragraph 1). In addition, *C. nigriceps* in particular can increase nest occurance because its pruning of the apial buds results in denser canopies which can provide further concealment of bird nests from predators (Discussion, paragraph 2).

d. Table of models\newline

| Model number | height_cm | species_name | Interaction?              |  
|--------------|:---------:|:------------:|:-------------------------:|
| 0            |  X        |      X       |  No interaction (+)       |
| 1            |  X        |      X       |  Yes Interaction (*)      | 


e. 


## Problem 3. Affective and exploratory visualizations

a. Comparing visualizations\newline

content

b. Designing an analysis\newline

One analysis I could apply to answer my research question of how my computer screen time compares on weekdays vs weekends is a t-test. This would be an appropriate test to answer my question as it would compare the average screen time on weekdays vs weekends. It makes sense for the predictor variable as weekday vs weekend are categorical predictor variables that can be used as the groups that I am comparing. It makes sense for the response variable of computer screen time as that is a continuous variable. 

c. Check any assumptions and run your analysis\newline

Checking for equal variances
```{r}

var.test



```





