---
title: "ENVS 193 Final Script"
author: "Jared Umphress"
subtitle: "June 9, 2026"
format: pdf

execute:
  warning: false
  message: false
---

Questions for Professor Bui:
- If I am comparing means would it be better to make a point range graph?


## Set Up\newline

[Link to GitHub repository](https://github.com/Jared363/ENVS-193DS_spring-2026_final)


Read in packages 
```{r}

library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(ggplot2)
library(ggeffects)
library(flextable)
library(MuMIn)
library(DHARMa)

```

Read in data
```{r}

# bringing in screen_time.csv (personal data) with read_csv function 
#   and storing it as object called my_data
my_data <- read_csv(here("data/screen_time.csv"))

# bringing in nest_data_final.csv with read_csv function and storing it 
#   as object called nest_data
nests <- read_csv(here("data/nest_data_final.csv"))

```

## Problem 1. Research writing

a. Transparent statistical methods\newline

In part 1 of the results section, they used a logistic regression. In part 2 of the results sections, they used a chi squared test.

b. Suggestions for rewriting\newline

Our results suggest that distance to water edge predicts the probability of American Avocet microhabitat use. In addition, we found evidence to support that there is a relationship between birds species (American Avocet, Foster's Tern, Black-necked Stilt) and habitat (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat).

c. Further analysis\newline

Algae concentration is another variable that could influence the probability of American Avocet microhabitat use as algae is another important primary producer in aquatic ecosystems that influences food availability. It is a continuous predictor variable measured in mg dry weight/L water, and it can be collected in 1 liter containers with the dry weight measured in lab. 

## Problem 2. Data analysis

a. Clean and wrangle\newline
```{r}

# cleaning nests data and storing it as nests_clean
nests_clean <- nests |>
  # selecting for just three columns
  select(height_cm, case_control, ant_species) |>
  # filtering for only three ant species
  filter(ant_species %in% c("RRB","AB","BBR")) |>
  # creating a new column to store full scientific names for ants
  #   based on abreviations
   mutate(species_name = case_match(
    ant_species,
    "RRB" ~ "Crematogaster mimosae", 
    "AB" ~ "Crematogaster sjostedti", 
    "BBR" ~ "Crematogaster nigriceps"
  )) |>
  # ordering the species level
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                       "AB",
                                       "RBR",
                                       "BBR"))

# display structure of nests_clean data
str(nests_clean)

# display 10 random rows from the nests_clean data
slice_sample(nests_clean, 
             n = 10, 
             by = NULL, 
             weight_by = NULL, 
             replace = FALSE)

```

b. Response variable\newline

For the nest occurence response variable, a 1 indicates that that a tree contained a nest, and 0 indicates that a tree was available for a nest. 

c. Purpose of study\newline

Tree height, one of the predictor variables, is a continuous variable. As tree height increases, the probability of nest occurrence would increase because the nests are further off the ground and harder to get to for predators which reduces nest predation, which can increase lifetime fitness in birds (Discussion, paragraph 1). In addition, taller trees likely have more branches and leaves which can create a denser canopy with or without the help of *C. nigriceps* that can provide further concealment of bird nests from predators (Discussion, paragraph 1).

Ant species, the other predictor variable, is a categorical variable with the four ant species representing four categories. The presence of *C. nigriceps* and *C. mimosae* would increase nest occurence as they are aggressive defenders of host trees and thus can reduce the risk of nest predation compared to *C. sjostedi* and *T. penzigi* which are less aggressive host tree defenders (Discussion, paragraph 1). In addition, *C. nigriceps* in particular can increase nest occurance because its pruning of the apial buds results in denser canopies which can provide further concealment of bird nests from predators (Discussion, paragraph 2).

d. Table of models\newline

| Model number | height_cm | species_name | Interaction?              |  
|--------------|:---------:|:------------:|:-------------------------:|
| 0            |  X        |      X       |  No interaction (+)       |
| 1            |  X        |      X       |  Yes Interaction (*)      | 


e. Run the models\newline

```{r}

# running a generalized linear model using glm 
#   and storing as nest_model_1 object
nest_model_1 <- glm(
                    # case_control as a response to height and species
                    #   with no interaction
                    case_control ~ height_cm + species_name,
                    # using nests_clean data
                    data = nests_clean,
                    # setting glm to binomial glm specifically
                    family = "binomial")

# display result of model 1
nest_model_1


nest_model_2 <- glm(
                    # case_control as a response to height and species
                    #   with interaction (using a * instead of +)
                    case_control ~ height_cm * species_name,
                    # using nests_clean data
                    data = nests_clean,
                    # setting glm to binomial glm specifically
                    family = "binomial")

# display result of model 2
nest_model_2

```

f. Select the best model\newline

```{r}

# AICc is Akaike's Information Criterion function
AICc(
  # running AIC for model 1 and model 2
  nest_model_1,
  nest_model_2
) |>
  # arranging the AIC scores by lowest at the top and highest at the bottom
  arrange(AICc)

```

The best model as determined by Akaike's Information Criterion (AIC) had case factor as the response variable with tree height and ant species as interacting predictor variables. 

g. Check the diagnostics\newline

```{r}
#| fig-height: 8
#| fig-width: 12


# checking diagnostics using simulated residuals function
plot(
  simulateResiduals(nest_model_2)
)

```

The points follow the QQ line relatively closely indicating that the residuals of the model follow a uniform distribution. The not significant dispersion result tells us that our residuals are not more dispersed than expected from the uniform distribution mentioned in the prior sentence. The not sigificant result for the outlier test tells us that there do not appear to be any outliers that could affect the model coefficients or predictions. The plot on the right shows us that the homoscedasticity, or variance of residuals overtime, stay relatively constant over time. All of these results tell us that the data fits the model assumptions and thus we can use a logistic regression to predict values for the response variable of nest probability. 

h. Visualize the model predictions\newline

```{r}

# model predictions
nest_predictions <- ggpredict(
  nest_model_2,   # model object
  terms = c("height_cm[all]", "species_name")   # predictor (in quotation marks)
) |>
  rename(species_name = group)

```

```{r}
# ggplot is base layer
ggplot(data = nests_clean, # using nests_clean data
       mapping = aes(
                     # x-axis is height
                     x = height_cm, 
                     # y axis is case_control (nest probability)
                     y = case_control, 
                     # coloring by species_name factor
                     color = as.factor(species_name))) +
  # changing ant colors manually
  scale_color_manual(values = c("Crematogaster sjostedti" = "skyblue",
                                "Crematogaster mimosae" = "orchid3",
                                "Crematogaster nigriceps" = "springgreen3")) +
  # changing ant fill colors manually
  scale_fill_manual(values = c("Crematogaster sjostedti" = "skyblue",
                                "Crematogaster mimosae" = "orchid3",
                                "Crematogaster nigriceps" = "springgreen3")) +
  # geom points is first layer 
  geom_point(
             mapping = aes(x = height_cm, # x-axis is acacia tree height
                           y = case_control, # y-axis is nest occurrence
                           color = species_name), # set to selected colors
             shape = 21, # change to open circle shape 
             alpha = 0.6 # changing transparency
            ) + 
  # second layer is geom line, showing predictions
  geom_line(data = nest_predictions, # data from model 2 prediction
            mapping = aes(x = x, # x-axis is height of acacia tree
                          y = predicted, # y axis is prediction 
                                         # of nest probability
                          color = species_name), # color by ant species
            linewidth = 1) +
  # third layer is geom_ribbon to show 95% confidence interval
  geom_ribbon(data = nest_predictions, # using data from model 2 prediction
              mapping = aes(x = x, # x-axis height of acacia
                            ymin = conf.low, # set low value
                            ymax = conf.high, # set high value
                            fill = species_name), # fill color by ant species
              inherit.aes = FALSE, # ignore global aes       
              alpha = 0.4) + # changing transparency
  # create a panel for each ant species
  facet_wrap(~ species_name) +   
  # relabeling x- and y-axis
  labs(x = "Height of Acacia Tree (cm)", 
       y = "Predicted Probability of Nest Presence", 
       color = "species_name", # color by ant species
       title = "Predicted Nest Presence by Ant Species and Acacia Height (cm)") +
  # changing theme from default
  theme_minimal() +
  theme(
        # removing panel background
        panel.background = element_rect(),
        # removing panel grid lines
        panel.grid = element_blank(),
        # removing legend 
        legend.position = "none") 

```

i. 










## Problem 3. Affective and exploratory visualizations

a. Comparing visualizations\newline

content

b. Designing an analysis\newline

One analysis I could apply to answer my research question of how my computer screen time compares on weekdays vs weekends is a t-test. This would be an appropriate test to answer my question as it would compare the average screen time on weekdays vs weekends. It makes sense for the predictor variable as weekday vs weekend are categorical predictor variables that can be used as the groups that I am comparing. It makes sense for the response variable of computer screen time as that is a continuous variable. 

c. Check any assumptions and run your analysis\newline

Checking for equal variances
```{r}

var.test



```





